---
title: "getSymbols and replaceSymbols"
author: "Daniel Lill"
date: "1 December 2018"
output: html_document
---

```{r setup, include=FALSE}
library(dMod)
library(dplyr)
library(stringr)
```

If, during day 2 you were wondering about some details of the underlying implementation of *insert*, one question could have been:

> How does insert() know, which parts of a symbolically defined function are parameters?

The answer is: it lets R *parse* the equations to find out.

Suppose, you have a Michaelis-Menten equation with log-transformed parameters.
```{r}
eqn <- c("P" = "exp(logVmax) * S / (exp(logKm) + S)")
```

If the equation is parsed, you will get a table of the parsed data:
```{r}
getParseData(parse(text = eqn)) %>% head(10)
```

## getSymbols and replaceSymbols
The function *getSymbols* now extracts the text of all elements with *token == "SYMBOL"*
```{r}
getSymbols(eqn)
```

Similarly, *replaceSymbols* replaces the text of these symbols
```{r}
replaceSymbols("logVmax", "VMAX", eqn)
```


## Safer than string-replacements
This approach is safer than string manipulations with regexes. Consider this simple case, where regexes are more error-prone.
```{r}
eqn <- c("k + k_1 + k.1")
replaceSymbols("k", "Santa", eqn)
```

This will fail
```{r}
str_replace_all(eqn, "k", "Santa")

```

This fails as well
```{r}
str_replace_all(eqn, "\\bk\\b", "Santa")
```











