---
title: "msParframe"
author: "Daniel Lill"
date: "12 December 2018"
output: html_document
---


```{r setup, include=FALSE}
library(tidyverse)

library(dMod)
```

# Enhacing reproducibility of mstrust

The multi-start optimization function `mstrust` can handle a `parframe` as input to the `pars`-argument.
This can be used to enhave the reproducibility of fitting results as will be shown later.

## Objective function
```{r}
truth <- c(Santa = 1, Rudolph = 2, Elf_Marcus = 3)
p <- setNames(nm = names(truth)) %>% 
  insert("x~exp(x)", x = .currentSymbols) %>% P
obj <- constraintL2(truth)
```

# Using mstrust twice - different parinits
```{r}
ms1 <- mstrust(obj, truth, fits = 10)
ms2 <- mstrust(obj, truth, fits = 10)
ms1 %>% map("parinit") %>% map(t) %>%  reduce(rbind)
ms2 %>% map("parinit") %>% map(t) %>%  reduce(rbind)
```

# Using msParframe
When passing a parframe, the samplefun and fits argument of mstrust are ignored, instead it will be looped over the different parameters right away.
Passing parframes is also helpful, if one wants to continue previous unconverged fits.

msParframe returns a parframe of sampled parameters, including the sampling center. This is helpful if the problem is difficult to optimize and one already has found a good fit. This way, it is guaranteed that the same optimum is found again.
Furthermore, it is possible to set a seed-argument to reproduce the results in a different R-session.
```{r}
mP <- msParframe(truth, n = 10, seed = 1)
ms3 <- mstrust(obj, mP, fits = 1000)

ms3 %>% map("parinit") %>% map(t) %>%  reduce(rbind)
```

# Continuing fits by passing parframes to mstrust
Let's simulate a difficult optimization problem by limiting the iterations
```{r}
mP <- msParframe(truth, n = 10)
ms3 <- mstrust(obj, mP, fits = 1000, iterlim = 2)
intermediate_pars <- ms3 %>% as.parframe()
print(intermediate_pars)
```

We can continue the optimization without problems
```{r}
ms3_continued <- mstrust(obj, intermediate_pars)
ms3_continued %>% as.parframe()
```

Happy mstrusting!

